---
title: "AMS"
author: "Elia Zanini, Marco Catoia, Alberto Polato"
date: "2026-02-23"
output: html_document
---

EXPLORATORY DATA ANALYSIS 
CHECKING VARIABLES IN THE DATASET, CHECKING AND DELETING NULL VALUES, CORRELATION ANALYSIS, ADDING NEW VARIABLES, ECC

```{r}

# Load libraries
library(tidyverse)
library(psych)
library(naniar)
library(car)
# Load dataset
df <- read.csv("diabetes.csv")

# Dimensioni
dim(df)

# Struttura
str(df)

# Nomi variabili
colnames(df)
```
creating diabetes variables and checking its proportion
```{r}
# Create binary diabetes variable
df <- df %>%
  mutate(diabetes = ifelse(glyhb > 7, 1, 0))

# Check prevalence
table(df$diabetes)
prop.table(table(df$diabetes))
```

missing values analysis
```{r}
# Count missing per variable
missing_summary <- sapply(df, function(x) sum(is.na(x)))
missing_percent <- sapply(df, function(x) mean(is.na(x)) * 100)

missing_table <- data.frame(
  Variable = names(missing_summary),
  Missing_Count = missing_summary,
  Missing_Percent = round(missing_percent, 2)
)

missing_table
```

deleting rows with a lot of null values (bp.2s and bp.2d) and all the rows with many null values
```{r}
df <- df %>%
  select(-bp.2s, -bp.2d)

df <- df[complete.cases(df), ]

```

renaming bp.1s and bp.1d and a clearer way
```{r}
df <- df %>%
  rename(
    systolic_bp = bp.1s,
    diastolic_bp = bp.1d
  )
```


```{r}
df
```

identifying numerical variables
```{r}
numeric_vars <- df %>%
  select(where(is.numeric)) %>%
  select(-id)  # Remove ID if numeric

# Descriptive statistics
describe(numeric_vars)
```

frequency of categorical variables
```{r}
# Gender
table(df$gender)
prop.table(table(df$gender))

# Location
table(df$location)
prop.table(table(df$location))

# Frame
table(df$frame)
prop.table(table(df$frame))

```
listing numerical variables
```{r}
library(dplyr)

numeric_df <- df %>%
  select(where(is.numeric)) %>%
  select(-id)   # togliamo identificatore

colnames(numeric_df)
```

computing correlation between numerical variables
```{r}
library(corrplot)

cor_matrix_spearman <- cor(numeric_df,
                           use = "pairwise.complete.obs",
                           method = "spearman")

round(cor_matrix_spearman, 2)


```
plotting the result
```{r}
library(corrplot)
corrplot(cor_matrix_spearman,
         method = "color",
         type = "upper",
         tl.cex = 0.7)
```


encoding categorical variables as factors to ensure regressions work properly
```{r}
# Convertire le variabili categoriche in fattori
df$location <- as.factor(df$location)
df$gender <- as.factor(df$gender)
df$frame <- as.factor(df$frame)

# Convertire la risposta in fattore
df$diabetes <- as.factor(df$diabetes)
str(df)
```


Trying to see the result of a first very simple logistic regression on the whole dataset: 
Logistic regression on all the variables before any transformation (no hypertension, BMI or WHR)
```{r}
mod <- glm(diabetes ~ chol + stab.glu + hdl + ratio + 
                       location + age + gender + height + weight + 
                       frame + systolic_bp + diastolic_bp + waist + hip + 
                       time.ppn,
           data = df,
           family = binomial)

# Riassunto del modello
summary(mod)
vif(mod)
```


Creating new variables BMI in kg/m^2 and WHR (wais hip ratio)
```{r}
df$BMI <- df$weight / (df$height^2) * 703
df$WHR <- df$waist / df$hip

df <- df %>%
  select(-weight, -height, -waist, -hip)
```

Creating variable hypertension (when systolic_bp >= 140 or diastolic_bp >= 90)
```{r}
df <- df %>%
  mutate(
    hypertension = ifelse(systolic_bp >= 140 | diastolic_bp >= 90, 1, 0)
  )
df$hypertension <- as.factor(df$hypertension)
str(df)
```

Trying a new simple logistic regression with the new variables WHR, BMI and hypertension
```{r}
mod2 <- glm(diabetes ~ chol + stab.glu + hdl +
                       location + age + gender + BMI + 
                       frame + WHR + hypertension + 
                       time.ppn,
           data = df,
           family = binomial)

#Riassunto del modello
summary(mod2)
vif(mod2)
```


Distribution of variables colored by diagnosis
```{r}
library(GGally)

numeric_df <- df[, sapply(df, is.numeric)]
numeric_df$id <- NULL

# assicuriamoci che diabetes sia presente per colorare
numeric_df$diabetes <- df$diabetes

# colonne da plottare (escludiamo diabetes e hypertension)
num_cols <- setdiff(names(numeric_df), c("diabetes", "hypertension"))

# funzione per correlazione Spearman overall
upper_cor <- function(data, mapping, ...) {
  x <- GGally::eval_data_col(data, mapping$x)
  y <- GGally::eval_data_col(data, mapping$y)
  r <- suppressWarnings(cor(x, y, use = "complete.obs", method = "spearman"))
  ggally_text(
    label = sprintf("%.2f", r),
    mapping = aes(),
    ...
  ) +
    theme_void()
}

suppressWarnings(
  ggpairs(
    numeric_df,
    columns = which(names(numeric_df) %in% num_cols),
    aes(color = factor(diabetes), alpha = 0.5),
    upper = list(continuous = upper_cor),
    progress = FALSE
  )
)

```
Bivariate distribution of variables to check the presence of subgroups or clusters
```{r}
pairs(df)
```


MCD for robust estimates and outliers detection (without time_ppn because it was giving computational errors)
```{r}
# selezione variabili continue fisiologiche
library(naniar)
library(MASS)
library(boot)
library(car)         # VIF
library(robustbase)  # robust regression
library(mclust)      # model-based clustering
library(broom)
library(sandwich)
library(lmtest)

library(robustbase)

cont_vars <- numeric_df[, c("BMI", "WHR", "stab.glu", "chol", "hdl",
                    "systolic_bp", "diastolic_bp",
                    "age")]

fit_MCD <- covMcd(x = cont_vars, alpha = 0.75)
fit_MCD


plot(fit_MCD, classic = TRUE)
```
tolerance elipsoide in 3d
```{r}
scatter3d(
  x = cont_vars$BMI,
  y = cont_vars$stab.glu,
  z = cont_vars$age,
  ellipsoid = TRUE
)
```


plot with threshold line
```{r}
md_rob <- sqrt(fit_MCD$mah)

plot(md_rob)
abline(h = sqrt(qchisq(0.975, df = ncol(cont_vars))), col = "red", lwd = 2)

```
theoretical threshold for identifying outliers
```{r}
p <- ncol(cont_vars)
cutoff <- sqrt(qchisq(0.975, df = p))  # soglia distanza robusta
cutoff
```
actual number of outliers and indices
```{r}
outliers <- md_rob > cutoff
table(outliers)
which(outliers)
```

plot with colored outliers 
```{r}
plot(md_rob)
abline(h = sqrt(qchisq(0.975, df = ncol(cont_vars))), col = "red", lwd = 2)
points(which(outliers), md_rob[outliers], col = "red", pch = 19)
```


removal of outliers and creating a new dataframe
```{r}
df_clean <- df[!outliers, ]
dim(df)        # dimensione originale
dim(df_clean)  # dimensione pulita
```
logistic regression on the clean dataset on all the variables after the transformations (BMI WHR hypertension)
```{r}
complete_logit_clean <- glm(diabetes ~ BMI + WHR + stab.glu + chol + hdl +
                     location + frame + age + ratio + hypertension + time.ppn,
                   data = df_clean,
                   family = binomial)
summary(complete_logit_clean)
library(car) 
vif(complete_logit_clean) 
```




logistic on the clean dataset (with transformed variables but without the categoricals frame location gender, and without ratio) 
```{r}
logit_clean <- glm(diabetes ~ BMI + WHR + stab.glu + chol + hdl +
                     hypertension + age + time.ppn,
                   data = df_clean,
                   family = binomial)
summary(logit_clean)
library(car) 
vif(logit_clean)  #metodo per verificare collinearità
```
robust logistic regression on the non-cleaned dataset on the already transformed variables (so BMI, WHR, hypertension)
```{r}
robust_logit_complete <- glmrob(diabetes ~ chol + stab.glu + hdl + ratio + 
                       location + age + gender + BMI + 
                       frame + WHR + hypertension + 
                       time.ppn,
                       data = df,
                       family = binomial)

summary(robust_logit_complete)
vif(robust_logit_complete)
```




robust logistic regression on the non-cleaned dataset (with transformed variables but without the categoricals frame location gender, and without ratio)
```{r}
library(robustbase)

robust_logit <- glmrob(diabetes ~ BMI + WHR + stab.glu + chol + hdl +
                         hypertension + age + time.ppn,
                       data = df,
                       family = binomial)

summary(robust_logit)
library(car) 
vif(robust_logit)  #metodo per verificare collinearità

```
Verifichiamo con plot e influence measures per i primi due modelli calcolati sul dataset completo se ci sono outliers o osservazioni particolarmente influenti che potrebbero peggiorare i nostri primi due modelli

Plot e influence measures primo modello (mod)
```{r}
plot(mod)
influence.measures(mod)

```

Plot e influence measures secondo modello (mod2)
```{r}
plot(mod2)
influence.measures(mod2)
```

Ora lo calcoliamo sul miglior modello sul dataset pulito per vedere se qualcosa è cambiato
```{r}
plot(logit_clean)
influence.measures(logit_clean)
```






Robust logistic regression with the most informative variables
```{r}
library(robustbase)
robust_logit2 <- glmrob(
  diabetes ~ BMI + stab.glu + hypertension + chol,
  data = df,
  family = binomial
)
summary(robust_logit2)
```
reimporting data and restructuring them as before
```{r}
  data_raw <- read.csv("diabetes.csv")
  data_raw$BMI <- data_raw$weight / (data_raw$height^2) * 703
  data_raw$WHR <- data_raw$waist / data_raw$hip
  data_raw$log_glyhb <- log(data_raw$glyhb)
  data_raw$bp.2s <- NULL
  data_raw$bp.2d <- NULL
  names(data_raw)[names(data_raw) == "bp.1s"] <- "systolic_bp"
  names(data_raw)[names(data_raw) == "bp.1d"] <- "diastolic_bp"
  data_raw$hypertension <- ifelse(data_raw$systolic_bp >= 140 | data_raw$diastolic_bp >= 90, 1, 0)
  data_raw <- na.omit(data_raw)
data_raw
```

Non-robust linear regression on all variables in the dataset (variables obtained from transformations are not included) on dataset with outliers
```{r}
lm1 <- lm(glyhb ~ chol + stab.glu + hdl + ratio + 
                       location + age + gender + height + weight + 
                       frame + systolic_bp + diastolic_bp + waist + hip + 
                       time.ppn,
                       data = data_raw)
summary(lm1)
```
non-robust linear regression (with hypertension, BMI, WHR) on dataset with outliers
```{r}
lm2 <- lm(glyhb ~ chol + stab.glu + hdl + ratio + 
                       location + age + gender + BMI + 
                       frame + WHR + hypertension + 
                       time.ppn,
                  data = data_raw)

summary(lm2)
```


Non-robust linear regression without categoricals and ratio on dataset with outliers
```{r}
lm3 <- lm(glyhb ~ BMI + WHR + stab.glu + hypertension +
               chol + hdl + age + time.ppn,
               data = data_raw)

summary(lm3)
```
outliers removal
```{r}
data_raw_clean <- data_raw[!outliers, ]
nrow(data_raw)
nrow(data_raw_clean)
```


Non-robust linear regression with all post-transformation vars on outlier-free dataset
```{r}
lm4 <- lm(glyhb ~ chol + stab.glu + hdl + ratio + 
                       location + age + gender + BMI + 
                       frame + WHR + hypertension + 
                       time.ppn,
               data = data_raw_clean)

summary(lm4)
```


Non-robust linear regression without categoricals and ratio on outlier-free dataset
```{r}
lm5 <- lm(glyhb ~ BMI + WHR + stab.glu + chol + hdl +
                  hypertension + age + time.ppn,
               data = data_raw_clean)

summary(lm5)
```


Robust linear regression on glyhb with all post-transformation variables
```{r}
library(MASS)
  model_rlm <- rlm(glyhb ~ chol + stab.glu + hdl + ratio + 
                       location + age + gender + BMI + 
                       frame + WHR + hypertension + 
                       time.ppn,
     data = data_raw)
  summary(model_rlm)
```
robust linear regression on glyhb without categoricals and ratio
```{r}
model_rlm2 <- rlm(glyhb ~ BMI + WHR + stab.glu + chol + hdl +
                  hypertension + age + time.ppn,
     data = data_raw)
  summary(model_rlm2)
```


DIAGNOSTICA MODELLI LOGISTICI E LINEARI E VALUTAZIONE PERFORMANCE

Influential Points, Cook's Distance and leverages for complete logit clean vs logit clean
Non robust models
```{r}
# Logistica pulita completa
infl_complete <- influence.measures(complete_logit_clean)

# Cook's distance
cook_complete <- cooks.distance(complete_logit_clean)
plot(cook_complete, type = "h",
     main = "Cook's Distance - complete logit clean",
     ylab = "Cook's D", xlab = "Osservazione")
abline(h = 4 / nrow(df_clean), col = "red", lty = 2)
text(which(cook_complete > 4 / nrow(df_clean)),
     cook_complete[cook_complete > 4 / nrow(df_clean)],
     labels = which(cook_complete > 4 / nrow(df_clean)),
     cex = 0.7, pos = 3)

# Leverage (hat values)
hat_complete <- hatvalues(complete_logit_clean)
p_complete   <- length(coef(complete_logit_clean))
n_complete   <- nrow(df_clean)
plot(hat_complete, type = "h",
     main = "Leverage – complete logit clean",
     ylab = "Hat values", xlab = "Osservazione")
abline(h = 2 * p_complete / n_complete, col = "red", lty = 2)

cat("Osservazioni con Cook's D elevato (complete logit clean):")
print(which(cook_complete > 4 / n_complete))

# Logistica pulita
cook_logit <- cooks.distance(logit_clean)
plot(cook_logit, type = "h",
     main = "Cook's Distance – logit clean",
     ylab = "Cook's D", xlab = "Osservazione")
abline(h = 4 / nrow(df_clean), col = "red", lty = 2)
text(which(cook_logit > 4 / nrow(df_clean)),
     cook_logit[cook_logit > 4 / nrow(df_clean)],
     labels = which(cook_logit > 4 / nrow(df_clean)),
     cex = 0.7, pos = 3)

cat("Osservazioni con Cook's D elevato (logit clean):")
print(which(cook_logit > 4 / nrow(df_clean)))
# Leverage (hat values) per logit_clean
hat_logit <- hatvalues(logit_clean)
p_logit   <- length(coef(logit_clean))
n_logit   <- nrow(df_clean)

plot(hat_logit, type = "h",
     main = "Leverage – logit clean",
     ylab = "Hat values", xlab = "Osservazione")

abline(h = 2 * p_logit / n_logit, col = "red", lty = 2)

cat("Osservazioni con leverage elevato (logit clean):")
print(which(hat_logit > 2 * p_logit / n_logit))
```


Pearson e devianza residua per la logistica

```{r}
# Residui di Pearson e di devianza per logit_clean
res_pearson  <- residuals(logit_clean, type = "pearson")
res_deviance <- residuals(logit_clean, type = "deviance")

par(mfrow = c(1, 2))
plot(fitted(logit_clean), res_pearson,
     main = "Residui di Pearson vs Fitted",
     xlab = "Valori adattati (prob)", ylab = "Residui Pearson")
abline(h = 0, col = "red")

plot(fitted(logit_clean), res_deviance,
     main = "Residui di Devianza vs Fitted",
     xlab = "Valori adattati (prob)", ylab = "Residui Devianza")
abline(h = 0, col = "red")
par(mfrow = c(1, 1))

#complete logit clean
#Residui di Pearson e di Devianza per complete_logit_clean
res_pearson_complete  <- residuals(complete_logit_clean, type = "pearson")
res_deviance_complete <- residuals(complete_logit_clean, type = "deviance")

par(mfrow = c(1, 2))

# Residui di Pearson vs valori adattati
plot(fitted(complete_logit_clean), res_pearson_complete,
     main = "Residui di Pearson vs Fitted (complete logit)",
     xlab = "Valori adattati (prob)", ylab = "Residui Pearson")
abline(h = 0, col = "red", lty = 2)

# Residui di Devianza vs valori adattati
plot(fitted(complete_logit_clean), res_deviance_complete,
     main = "Residui di Devianza vs Fitted (complete logit)",
     xlab = "Valori adattati (prob)", ylab = "Residui Devianza")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

Goodness of fit – Test di Hosmer-Lemeshow

```{r}
library(ResourceSelection)

# Hosmer-Lemeshow per i modelli logistici principali
hl_complete <- hoslem.test(
  as.numeric(as.character(df_clean$diabetes)),
  fitted(complete_logit_clean), g = 10)
cat("Hosmer-Lemeshow: complete_logit_clean")
print(hl_complete)

hl_logit <- hoslem.test(
  as.numeric(as.character(df_clean$diabetes)),
  fitted(logit_clean), g = 10)
cat("\n=== Hosmer-Lemeshow: logit_clean ===\n")
print(hl_logit)
```

Curva ROC e AUC non robust logit

```{r}
library(pROC)

# Probabilità predette
pred_complete <- predict(complete_logit_clean, type = "response")
pred_logit    <- predict(logit_clean, type = "response")
true_y        <- as.numeric(as.character(df_clean$diabetes))

roc_complete <- roc(true_y, pred_complete, quiet = TRUE)
roc_logit    <- roc(true_y, pred_logit,    quiet = TRUE)

# Plot curva ROC
plot(roc_complete, col = "steelblue", lwd = 2,
     main = "ROC curves - non-robust models")
lines(roc_logit, col = "darkorange", lwd = 2)
legend("bottomright",
       legend = c(paste0("complete_logit_clean (AUC=", round(auc(roc_complete), 3), ")"),
                  paste0("logit_clean (AUC=",          round(auc(roc_logit),    3), ")")),
       col = c("steelblue", "darkorange"), lwd = 2)

cat("AUC complete_logit_clean:", round(auc(roc_complete), 4), "\n")
cat("AUC logit_clean:         ", round(auc(roc_logit),    4), "\n")
```
robust logit
```{r}
library(pROC)

# Probabilità predette
pred_robust_complete <- predict(robust_logit_complete, type = "response")
pred_robust_logit    <- predict(robust_logit, type = "response")

# Variabili risposta (devono venire dai dataset usati nei modelli)
true_complete <- as.numeric(as.character(df$diabetes))
true_clean    <- as.numeric(as.character(df$diabetes))

# ROC
roc_robust_complete <- roc(true_complete, pred_robust_complete, quiet = TRUE)
roc_robust_logit    <- roc(true_clean, pred_robust_logit, quiet = TRUE)

# Plot
plot(roc_robust_complete, col = "darkgreen", lwd = 2,
     main = "ROC curves - Robust Models")

lines(roc_robust_logit, col = "purple", lwd = 2)

legend("bottomright",
       legend = c(
         paste0("robust_logit_complete (AUC=", round(auc(roc_robust_complete), 3), ")"),
         paste0("robust_logit (AUC=", round(auc(roc_robust_logit), 3), ")")
       ),
       col = c("darkgreen", "purple"),
       lwd = 2)

# AUC
cat("AUC robust_complete:", round(auc(roc_robust_complete), 4), "\n")
cat("AUC robust_logit:   ", round(auc(roc_robust_logit), 4), "\n")
```
Confusion matrices
```{r}
# funzione per creare confusion matrix
get_cm <- function(pred_prob, true_y, threshold = 0.5) {
  pred_class <- ifelse(pred_prob >= threshold, 1, 0)
  table(Predicted = pred_class, Actual = true_y)
}

# variabili risposta
true_clean <- as.numeric(as.character(df_clean$diabetes))
true_df    <- as.numeric(as.character(df$diabetes))

# confusion matrices
cm_complete <- get_cm(pred_complete, true_clean)
cm_logit    <- get_cm(pred_logit, true_clean)
cm_rcomplete <- get_cm(pred_robust_complete, true_df)
cm_rlogit    <- get_cm(pred_robust_logit, true_df)

# stampa con nomi dei modelli
cat("\nConfusion Matrix – complete_logit_clean\n")
print(cm_complete)

cat("\nConfusion Matrix – logit_clean\n")
print(cm_logit)

cat("\nConfusion Matrix – robust_logit_complete\n")
print(cm_rcomplete)

cat("\nConfusion Matrix – robust_logit\n")
print(cm_rlogit)
```

```{r}
# funzione per confusion matrix
get_cm <- function(pred_prob, true_y, threshold = 0.5) {
  pred_class <- ifelse(pred_prob >= threshold, 1, 0)
  table(Predicted = pred_class, Actual = true_y)
}

# funzione per metriche
metrics_from_cm <- function(cm, model_name){
  
  TP <- cm["1","1"]
  TN <- cm["0","0"]
  FP <- cm["1","0"]
  FN <- cm["0","1"]
  
  accuracy  <- (TP + TN) / (TP + TN + FP + FN)
  precision <- TP / (TP + FP)
  recall    <- TP / (TP + FN)
  specificity <- TN / (TN + FP)
  f1 <- 2 * (precision * recall) / (precision + recall)
  
  data.frame(
    Model = model_name,
    Accuracy = accuracy,
    Precision = precision,
    Recall = recall,
    Specificity = specificity,
    F1_score = f1
  )
}

# variabili risposta
true_clean <- as.numeric(as.character(df_clean$diabetes))
true_df    <- as.numeric(as.character(df$diabetes))

# confusion matrices
cm_complete <- get_cm(pred_complete, true_clean)
cm_logit    <- get_cm(pred_logit, true_clean)
cm_rcomplete <- get_cm(pred_robust_complete, true_df)
cm_rlogit    <- get_cm(pred_robust_logit, true_df)

# metriche
m1 <- metrics_from_cm(cm_complete, "complete_logit_clean")
m2 <- metrics_from_cm(cm_logit, "logit_clean")
m3 <- metrics_from_cm(cm_rcomplete, "robust_logit_complete")
m4 <- metrics_from_cm(cm_rlogit, "robust_logit")

# tabella finale
results <- rbind(m1, m2, m3, m4)

print(results)
```



Brier Score
Modelli non robusti
```{r}
brier <- function(pred_prob, true_y, model_name) {
  bs <- mean((pred_prob - true_y)^2)
  cat(sprintf("Brier Score (%s): %.4f\n", model_name, bs))
  # 0 = perfetto, 0.25 = modello nullo (baseline 50/50)
}

brier(pred_complete, true_y, "complete_logit_clean")
brier(pred_logit,    true_y, "logit_clean")
```
Modelli robusti
```{r}
# Probabilità predette dai modelli robusti
pred_robust_complete <- predict(robust_logit_complete, type = "response")
pred_robust_logit    <- predict(robust_logit, type = "response")

# Variabile risposta dai rispettivi dataset
true_y_complete <- as.numeric(as.character(df_clean$diabetes))  # usa il dataset corretto del modello
true_y_logit    <- as.numeric(as.character(df_clean$diabetes))

# Calcolo Brier score
brier(pred_robust_complete, true_y_complete, "robust_logit_complete")
brier(pred_robust_logit,    true_y_logit,    "robust_logit")
```




cross entropy
Non robust models
```{r}
log_loss <- function(pred_prob, true_y, model_name) {
  eps <- 1e-15
  pred_prob <- pmax(pmin(pred_prob, 1 - eps), eps)
  ll <- -mean(true_y * log(pred_prob) + (1 - true_y) * log(1 - pred_prob))
  cat(sprintf("Log-Loss (%s): %.4f\n", model_name, ll))
}

log_loss(pred_complete, true_y, "complete_logit_clean")
log_loss(pred_logit,    true_y, "logit_clean")
```
Robust models
```{r}
log_loss(pred_robust_complete, true_y, "robust_logit_complete")
log_loss(pred_robust_logit,    true_y, "robust_logit")
```




Omoschedasticità, test di Breusch-pagan

```{r}
library(lmtest)
library(sandwich)

for (mod_name in c("lm1", "lm2", "lm3", "lm4", "lm5", "model_rlm", "model_rlm2")) {
  mod_obj <- get(mod_name)
  bp <- bptest(mod_obj)
  cat(sprintf("\n=== Breusch-Pagan test: %s ===\n", mod_name))
  print(bp)
}

# Grafico scale location
par(mfrow = c(1, 2))

plot(lm1, which = 3, main = "lm1")
plot(lm2, which = 3, main = "lm2")
plot(lm3, which = 3, main = "lm3")
plot(lm4, which = 3, main = "lm4")
plot(lm5, which = 3, main = "lm5")
plot(model_rlm, which = 3, main = "rlm1")
plot(model_rlm2, which = 3, main = "rlm2")

par(mfrow = c(1, 1))
```


Influential Points per linear mdoels non robusti

```{r}
diagnose_lm <- function(model, model_name) {
  n <- nrow(model$model)
  p <- length(coef(model))

  cook  <- cooks.distance(model)
  hat   <- hatvalues(model)
  dffit <- dffits(model)

  cat(sprintf("\n=== Influential Points: %s ===\n", model_name))
  cat("Soglia Cook's D   : 4/n =", round(4 / n, 4), "\n")
  cat("Soglia Leverage   : 2p/n =", round(2 * p / n, 4), "\n")
  cat("Soglia DFFITS     : 2*sqrt(p/n) =", round(2 * sqrt(p / n), 4), "\n")
  cat("Obs. con Cook > soglia:", sum(cook > 4 / n), "\n")
  cat("Obs. con hat > soglia :", sum(hat  > 2 * p / n), "\n")
  cat("Obs. con |DFFITS| > soglia:", sum(abs(dffit) > 2 * sqrt(p / n)), "\n")

  # Plot Cook
  plot(cook, type = "h",
       main = paste("Cook's Distance:", model_name),
       ylab = "Cook's D", xlab = "Osservazione")
  abline(h = 4 / n, col = "red", lty = 2)
}

diagnose_lm(lm1, "lm1")
diagnose_lm(lm2, "lm2")
diagnose_lm(lm3, "lm3")
diagnose_lm(lm4, "lm4")
diagnose_lm(lm5, "lm5")

```

Residuals vs Fitted, qq, scale location, leverage)

```{r}
par(mfrow = c(2, 2))
plot(lm1, main = "lm1")
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))
plot(lm2, main = "lm2")
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))
plot(lm3, main = "lm3")
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))
plot(lm4, main = "lm4")
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))
plot(lm5, main = "lm5")
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))
plot(model_rlm, main = "model_rlm")
par(mfrow = c(1, 1))

par(mfrow = c(2, 2))
plot(model_rlm2, main = "model_rlm2")
par(mfrow = c(1, 1))
```

VIF per modelli lineari

```{r}
library(car)

for (mod_name in c("lm1", "lm2", "lm3", "lm4", "lm5", "model_rlm", "model_rlm2")) {
  cat(sprintf("\n=== VIF: %s ===\n", mod_name))
  print(vif(get(mod_name)))
}
```

AIC, BIC e R² a confronto tra modelli lineari

```{r aic-bic-linear}
model_comparison_lm <- data.frame(
  Model     = c("lm1", "lm2", "lm3", "lm4", "lm5",
                  "model_rlm", "model_rlm2"),
  AIC         = c(AIC(lm1), AIC(lm2), AIC(lm3), AIC(lm4), AIC(lm5),
                  AIC(model_rlm), AIC(model_rlm2)),
  BIC         = c(BIC(lm1), BIC(lm2), BIC(lm3), BIC(lm4), BIC(lm5),
                  BIC(model_rlm), BIC(model_rlm2)),
  Adj_R2      = c(summary(lm1)$adj.r.squared,
                  summary(lm2)$adj.r.squared,
                  summary(lm3)$adj.r.squared,
                  summary(lm4)$adj.r.squared,
                  summary(lm5)$adj.r.squared,
                  NA, NA)  # rlm non ha R² standard
)
model_comparison_lm <- model_comparison_lm[order(model_comparison_lm$AIC), ]
print(model_comparison_lm)
```





RANDOM FOREST
```{r}
library(randomForest)
library(pROC)

# training sul dataset pulito
set.seed(42)
rf_model <- randomForest(
  diabetes ~ BMI + WHR + stab.glu + chol + hdl + hypertension + age + time.ppn,
  data       = df,
  ntree      = 500,
  mtry       = 3,        # default per classificazione: floor(sqrt(p))
  importance = TRUE,
  keep.forest = TRUE
)

print(rf_model)
# OOB error rate = stima out-of-bag dell'errore, non serve un test set separato
# Cerca il mtry ottimale minimizzando OOB error
set.seed(42)
tune_rf <- tuneRF(
  x          = df[, c("BMI", "WHR", "stab.glu", "chol", "hdl", "age", "time.ppn")],
  y          = df$diabetes,
  mtryStart  = 2,
  stepFactor = 1.5,
  improve    = 0.01,
  ntreeTry   = 300,
  trace      = FALSE,
  plot       = TRUE
)
best_mtry <- tune_rf[which.min(tune_rf[, "OOBError"]), "mtry"]
cat("Miglior mtry:", best_mtry, "\n")

# Riallena con il mtry ottimale
set.seed(42)
rf_final <- randomForest(
  diabetes ~ BMI + WHR + stab.glu + chol + hdl + hypertension + age + time.ppn,
  data        = df,
  ntree       = 500,
  mtry        = best_mtry,
  importance  = TRUE,
  keep.forest = TRUE
)
print(rf_final)
# MeanDecreaseAccuracy: quanto peggiora l'OOB accuracy senza quella variabile
# MeanDecreaseGini: impurità media di Gini (meno interpretabile ma più stabile)
importance(rf_final)
varImpPlot(rf_final,
           main = "Variable Importance – Random Forest",
           col  = "steelblue", pch = 16)
# Probabilità predette OOB
pred_rf_prob <- rf_final$votes[, 2]   # prob. diabetico
true_y       <- as.numeric(as.character(df$diabetes))

roc_rf <- roc(true_y, pred_rf_prob, quiet = TRUE)
cat("AUC Random Forest:", round(auc(roc_rf), 4), "\n")

# Probabilità predette modelli robusti
pred_robust_complete <- predict(robust_logit_complete, type = "response")
pred_robust_logit    <- predict(robust_logit,          type = "response")

roc_robust_complete <- roc(true_y, pred_robust_complete, quiet = TRUE)
roc_robust_logit    <- roc(true_y, pred_robust_logit,    quiet = TRUE)

# Confronto curva ROC: RF vs Robusti
plot(roc_rf,             col = "forestgreen", lwd = 2,
     main = "Confronto ROC: RF vs Modelli Robusti")
lines(roc_robust_complete, col = "blue", lwd = 2)
lines(roc_robust_logit,    col = "purple",    lwd = 2)
legend("bottomright",
       legend = c(
         paste0("Random Forest (AUC=",           round(auc(roc_rf),             3), ")"),
         paste0("robust_logit_complete (AUC=",   round(auc(roc_robust_complete),3), ")"),
         paste0("robust_logit (AUC=",            round(auc(roc_robust_logit),   3), ")")
       ),
       col = c("forestgreen", "blue", "purple"), lwd = 2)


```



BOOTSTRAP SUI MODELLI CHE HANNO OTTENUTO RISULTATI MIGLIORI
bootstrap per la stabilità dei coefficienti
```{r}
library(boot)
boot_logit <- function(data, indices) {
  d <- data[indices, ]
  fit <- glm(diabetes ~ BMI + stab.glu + hypertension + chol + age,
             data = d, family = binomial)
  coef(fit)
  }

set.seed(42)
boot_res <- boot(data = df_clean, statistic = boot_logit, R = 1000)
boot.ci(boot_res, type = "perc", index = 2)  #BMI
boot.ci(boot_res, type = "perc", index = 3)  #stab.glu
boot.ci(boot_res, type = "perc", index = 4)  #hyper
boot.ci(boot_res, type = "perc", index = 5)  #chol
boot.ci(boot_res, type = "perc", index = 6)  #age


par(mfrow = c(2, 3))
for (i in 2:6) {
  hist(boot_res$t[, i], breaks = 40, col = "steelblue", border = "white",
       main = names(coef(glm(diabetes ~ BMI + stab.glu + hypertension + chol + age,
                              data = df_clean, family = binomial)))[i],
       xlab = "Valore coefficiente", ylab = "Densità", probability = TRUE)
  abline(v = boot_res$t0[i], col = "red", lwd = 2, lty = 2)
  abline(v = 0, col = "gray50", lwd = 1)
}
par(mfrow = c(1, 1))

# Tabella riassuntiva: stima, bias, SE bootstrap
print(boot_res)

```






Modello bayesiano con assunzioni sul prior dei coefficienti
bayesian model with hamiltonian monte carlo
```{r}
library(brms)

bayes_model <- brm(
  diabetes ~ BMI + stab.glu + age + (1 | location),
  data = df_clean, family = bernoulli(),
  prior = c(prior(normal(0, 2.5), class = "b")),
  chains = 4, iter = 2000
)
plot(bayes_model)
summary(bayes_model)

```






UNSUPERVISED LEARNING - MODEL BASED CLUSTERING E K-MEANS
UNSUPERVISED LEARNING - MODEL BASED CLUSTERING E K-MEANS

```{r}
library(tclust)
library(mclust)
library(cluster)

# Variabili per clustering
X <- df[, c("BMI", "WHR", "stab.glu", "chol", "hdl", "age")]
X_scaled <- scale(X)

# true_diabetes su df
true_diabetes <- as.numeric(as.character(df$diabetes))
```

mbc con valutazione bic
```{r}
# Model Based Clustering
model <- Mclust(X_scaled, G = 2)
plot(model, what = "classification")
plot(model, what = "BIC")
summary(model)

df$cluster_mclust <- as.factor(model$classification)

cat("Mclust: non diabetici vs diabetici\n")
tab_mclust <- table(Cluster = df$cluster_mclust, Diabetes = true_diabetes)
print(tab_mclust)
print(prop.table(tab_mclust, margin = 1))

ari_mclust <- adjustedRandIndex(model$classification, true_diabetes)
cat(sprintf("\nARI Mclust (K=%d) vs diabetes: %.4f\n",
            length(unique(model$classification)), ari_mclust))
```


elbow plot e silhouette
```{r}
# Elbow plot
wss <- sapply(1:8, function(k) {
  kmeans(X_scaled, centers = k, nstart = 25, iter.max = 100)$tot.withinss
})
plot(1:8, wss, type = "b", pch = 19, col = "steelblue",
     main = "Elbow Plot",
     xlab = "Numero di cluster K", ylab = "WSS totale")
abline(v = 2, col = "red", lty = 2)

# Silhouette
avg_sil <- sapply(2:8, function(k) {
  km  <- kmeans(X_scaled, centers = k, nstart = 25, iter.max = 100)
  sil <- silhouette(km$cluster, dist(X_scaled))
  mean(sil[, 3])
})
plot(2:8, avg_sil, type = "b", pch = 19, col = "steelblue",
     main = "Silhouette Media",
     xlab = "Numero di cluster K", ylab = "Silhouette media")
abline(v = which.max(avg_sil) + 1, col = "red", lty = 2)
cat("K ottimale:", which.max(avg_sil) + 1, "\n")
```

kmeans
```{r}
# K-Means con K=2
set.seed(42)
km2 <- kmeans(X_scaled, centers = 2, nstart = 25, iter.max = 100)

ari_km2 <- adjustedRandIndex(km2$cluster, true_diabetes)
cat(sprintf("ARI K-Means (K=2): %.4f\n", ari_km2))

tab_km2 <- table(KMeans = km2$cluster, Diabetes = true_diabetes)
print(tab_km2)
print(prop.table(tab_km2, margin = 1))
```


robust clustering
```{r}
set.seed(42)
fit_trimmed_k_means <- tclust(
  x            = X_scaled,
  k            = 2,
  alpha        = 0.05,
  nstart       = 50,
  restr.fact   = 5,
  equal.weights = FALSE
)

mask      <- fit_trimmed_k_means$cluster != 0
ari_tclust <- adjustedRandIndex(fit_trimmed_k_means$cluster[mask], true_diabetes[mask])
cat(sprintf("ARI tclust (K=2, alpha=0.05): %.4f\n", ari_tclust))
cat("Trimmed observations:", sum(!mask), "\n")

tab_tclust <- table(TCLUST = fit_trimmed_k_means$cluster[mask], Diabetes = true_diabetes[mask])
print(tab_tclust)
print(prop.table(tab_tclust, margin = 1))

# Confronto Mclust vs tclust
table(Mclust = model$classification[mask],
      TCLUST = fit_trimmed_k_means$cluster[mask])
```

pca
```{r}
pca     <- prcomp(X_scaled, center = FALSE, scale. = FALSE)
pc1     <- pca$x[, 1]
pc2     <- pca$x[, 2]
var_exp <- round(summary(pca)$importance[2, 1:2] * 100, 1)

# Colori k-means
col_cluster  <- c("steelblue", "tomato")[km2$cluster]
# Colori diagnosi reale
col_diabetes <- c("steelblue", "tomato")[true_diabetes + 1]
# Colori tclust (0 = trimmed = grigio)
col_tclust <- c("gray70", "steelblue", "tomato")[fit_trimmed_k_means$cluster + 1]
#Colori mclust
col_mclust <- c("steelblue", "tomato")[model$classification]
```

visualizzazione mbc vs diagnosi reale
```{r}
par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))

plot(pc1, pc2, col = col_mclust, pch = 16, cex = 0.8,
     main = "Groups found by MBC",
     xlab = paste0("PC1 (", var_exp[1], "% var)"),
     ylab = paste0("PC2 (", var_exp[2], "% var)"))

plot(pc1, pc2, col = col_diabetes, pch = 16, cex = 0.8,
     main = "Real diagnosis of diabetes",
     xlab = paste0("PC1 (", var_exp[1], "% var)"),
     ylab = paste0("PC2 (", var_exp[2], "% var)"))

par(mar = c(0, 0, 0, 0))
plot.new()
legend(x = 0.2, y = 0.8,
       legend = c("Cluster 1", "Cluster 2"),
       col    = c("steelblue", "tomato"),
       pch    = 16, bty = "n", cex = 1.1,
       title  = "MBC")
legend(x = 0.6, y = 0.8,
       legend = c("Non-diabetic", "Diabetic"),
       col    = c("steelblue", "tomato"),
       pch    = 16, bty = "n", cex = 1.1,
       title  = "Real diagnosis")


par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
```


visualizzazione kmeans vs diagnosi reale
```{r}
par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))

plot(pc1, pc2, col = col_cluster, pch = 16, cex = 0.8,
     main = "Groups fopund by K-Means",
     xlab = paste0("PC1 (", var_exp[1], "% var)"),
     ylab = paste0("PC2 (", var_exp[2], "% var)"))

plot(pc1, pc2, col = col_diabetes, pch = 16, cex = 0.8,
     main = "Real diagnosis of diabetes",
     xlab = paste0("PC1 (", var_exp[1], "% var)"),
     ylab = paste0("PC2 (", var_exp[2], "% var)"))

par(mar = c(0, 0, 0, 0))
plot.new()
legend(x = 0.2, y = 0.8,
       legend = c("Cluster 1", "Cluster 2"),
       col    = c("steelblue", "tomato"),
       pch    = 16, bty = "n", cex = 1.1,
       title  = "K-Means group")
legend(x = 0.6, y = 0.8,
       legend = c("Non diabetic", "Diabetic"),
       col    = c("steelblue", "tomato"),
       pch    = 16, bty = "n", cex = 1.1,
       title  = "Real diagnosis")

par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
```

visualizzazione tclust vs diagnosi reale
```{r}
par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))

plot(pc1, pc2, col = col_tclust, pch = 16, cex = 0.8,
     main = "Tclust groups",
     xlab = paste0("PC1 (", var_exp[1], "% var)"),
     ylab = paste0("PC2 (", var_exp[2], "% var)"))

plot(pc1, pc2, col = col_diabetes, pch = 16, cex = 0.8,
     main = "Real diagnosis",
     xlab = paste0("PC1 (", var_exp[1], "% var)"),
     ylab = paste0("PC2 (", var_exp[2], "% var)"))

par(mar = c(0, 0, 0, 0))
plot.new()
legend(x = 0.1, y = 0.8,
       legend = c("Cluster 1", "Cluster 2", "Trimmed"),
       col    = c("steelblue", "tomato", "gray70"),
       pch    = 16, bty = "n", cex = 1.1,
       title  = "Tclust group")
legend(x = 0.6, y = 0.8,
       legend = c("Non diabetic", "Diabetic"),
       col    = c("steelblue", "tomato"),
       pch    = 16, bty = "n", cex = 1.1,
       title  = "Real diagnosis")

par(mfrow = c(1, 1), mar = c(5, 4, 4, 2))
```

ari a confronto
```{r}
cat("ARI\n")
cat(sprintf("Mclust  (K=%d)          : ARI = %.4f\n",
            length(unique(model$classification)), ari_mclust))
cat(sprintf("K-Means (K=2)           : ARI = %.4f\n", ari_km2))
cat(sprintf("tclust  (K=2, a=0.05)   : ARI = %.4f\n", ari_tclust))
```

